Analytic Occultation Light Curves in Reflected Light

### 2020-05-06 tex/ms.tex
more text

As in \citet{Luger2019}, we compute fluxes by first expanding the surface
in terms of spherical harmonics.
While in \citet{Luger2019} we expanded the emissivity of the surface, here
we instead expand the spherical albedo $A$.
Specifically, if $\bvec{y}$ is the vector
of spherical harmonic coefficients describing the albedo anywhere on
the surface and
$\by$ is the spherical harmonics basis (Equation~\ref{eq:by}),
the albedo $A$ at a point $(x, y)$ on the sky-projected disk of the body
is given by the dot product
%
\begin{align}
    \label{eq:albedo}
    A(x, y) = \by^\top (x, y) \, \bvec{y}
    \quad.
\end{align}
%
The flux measured from this body is the surface integral over the visible
portion of the body's projected disk of the albedo $A$
times the illumination profile $\mathcal{I}$ of the surface, given by
Lambert's law as
%
\begin{align}
    \label{eq:LambertsLaw}
    \mathcal{I}(\phi) = \mathcal{I}_0 \, \text{max}\big( 0, \cos\phi \big)
    \quad,
\end{align}
%
where $\mathcal{I}_0$ is the peak illumination (at the sub-source point).
%

@@

Our last sample light curve is Figure~\ref{fig:earthsun}, which shows
a secondary eclipse light curve of the Earth as it is occulted by the
Sun. The model for the Earth is the same as above, and the observer is
now close to the ecliptic, but slightly misaligned so that the
Earth is occulted behind a solar latitude of $30^\circ$ (i.e., at
a solar impact parameter of $0.5$). The observation takes place at the
June solstice, so the Earth is tilted by $23.5^\circ$ out of the page.
As before, the top panel shows the light curve in thermal light; this
is similar to the top panel of Figure~13 in \citet{Luger2019}. The orange
dots again correspond to the numerical solution.
%
The center panel shows the same light curve in reflected light.
Because the observation occurs very close to full phase, the normalized
light curves look very similar to each other.
The bottom panel shows the difference between the two
(reflected minus thermal), which is only on the order of a few percent.
%
In fact, because
the illumination profile is proportional to the cosine of the viewing
angle, $\mu$, and the reflection is assumed to be isotropic, the
(normalized) secondary eclipse light curve in reflected light is to
good approximation equal
to a limb-darkened thermal occultation light curve with
linear limb darkening coefficient $u_1 = 1$. As we will see later, for very
close-in planets this approximation breaks down, since the illumination
phases at secondary eclipse ingress and egress are sufficiently different
from full phase.

@@

% DEBUG
\clearpage

@@

\subsection{Extended illumination source}
\label{sec:extended}
Planets to consider: 55 Cnc e, Kelt-9b.
Assuming $R_\star > R_p$, the angular extent of the terminator
past $\nicefrac{\pi}{2}$ is
%
\begin{proof}{}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{R_\star - R_p}{d} \right)
\end{proof}
%
where $R_\star$ is the radius of the star,
$R_p$ is the radius of the planet, and $d$ is the separation of their
centers.

For Kelt-9b, $\nicefrac{R_p}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$, where $a$ is the semi-major axis of the
planet \citep{Wong2019}. Asssuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020}, we find that
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of the planet. Figure~\ref{fig:extended} shows...

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/extended.pdf}
        \oscaption{extended}{%
            Normalized surface intensity on Kelt-9b viewed in a
            Mollweide projection assuming a point illumination source
            (left) and accounting for the finite extent of the star
            (right). The day/night terminator extends about $17^\circ$
            past where it is in the point source case. The sub-stellar
            intensity is higher in the extended source case because the
            sub-planet point on the star is
            closer to the planet than in the case where the star is a point
            source located at the center of the star.
            \label{fig:extended}
        }
    \end{centering}
\end{figure}

Figure~\ref{fig:kelt9b} shows...

@@

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/kelt9b_eclipse.pdf}
        \oscaption{kelt9b}{%
            Normalized reflected light secondary eclipse model for Kelt-9b,
            assuming a spherical albedo $A = 0.2$. Blue is the \starry model
            assuming a point source illumination; purple accounts for
            the finite size of the star. The orange curves show
            approximate models computed using the \citet{MandelAgol2002}
            model: a sphere of uniform intensity (solid orange)
            and a sphere whose intensity falls as the cosine of the
            viewing angle from the center of the planet disk
            (dashed orange).
            Because Kelt-9b is so close to its host star, the illumination
            phase changes significantly from ingress to egress, and neither
            approximation accurately captures the behavior of the light curve.
            On the other hand, the point source approximation agrees well with
            the extended source solution, modulo an overall amplitude
            difference (see Figure~\ref{fig:kelt9b}).
            \label{fig:kelt9b_eclipse}
        }
    \end{centering}
\end{figure}

\subsection{Non-Lambertian scatterers}
\label{sec:nonlambertian}
Just change $\mathbf{I}$. \citep{OrenNayar1994}.

@@

\section{Applications}
\label{sec:applications}

@@

\subsection{Information}
\label{sec:information}

@@

Second, we must explicitly model the illumination of the body. We assume the
body is illuminated by a point-like source whose flux measured by the observer
is unity. In this case, the observed intensity at any
point on the surface is proportional to the cosine of the angle $\phi$ between
the incident light and the surface normal. Points for which
$\phi \ge \nicefrac{\pi}{2}$ are unilluminated and
therefore have an intensity of zero.
%
If the point-like illumination source is placed at sky coordinates
$(x_s, y_s, z_s)$ in units of the radius of the illuminated body,
the day/night terminator on the body is a half-ellipse
of semi-major axis unity that is fully described by its (signed) semi-minor
axis,
%
\begin{proof}{}
    \label{eq:b}
    b = -\frac{z_s}{r_s}
    \quad,
\end{proof}
%
where $r_s = \sqrt{x_s^2 + y_s^2 + z_s^2}$ is the distance to the source,
%
and the angle $\theta$ by which its semi-major axis is rotated away from the
$+x$-axis (a frame-dependent quantity, defined below).
Given this formulation, and assuming that
$r_s \gg 1$,
it is straightforward to show that the illumination
$\mathcal{I}$ at a point $(x, y)$ on the projected disk of the body is given
by the function
%
\begin{proof}{}
    \label{eq:illum}
    \mathcal{I}(b, \theta, r_s; x, y)&=
    \text{max}\bigg( 0, I(b, \theta, r_s; x, y) \bigg)
\end{proof}
%
where
%
\begin{proof}{}
    \label{eq:illum_poly}
    I(b, \theta, r_s; x, y) &= \frac{1}{\pi r_s^2}
    \bigg(
    -b_c\sin\theta x + b_c\cos\theta y - bz(x, y)
    \bigg)
\end{proof}
%
with $b_c \equiv \sqrt{1 - b^2}$ and $z(x, y) = \sqrt{1 - x^2 - y^2}$.
A quick note on the normalization of Equation~(\ref{eq:illum_poly}):
if we place the illumination source along the $+z$-axis at $(0, 0, 1)$,
the body is seen at full phase, so $b = -1$, $b_c = 0$, and
%
\begin{proof}{}
    \mathcal{I}_\text{full}(x, y) = \frac{\sqrt{1 - x^2 - y^2}}{\pi}
    \quad.
\end{proof}
%
Integrating this over the full disk, we obtain the reflected flux measured
by the observer if the body has unit spherical albedo,
%
\begin{proof}{}
    \mathcal{f}_\text{full} &=
    \int_{-1}^{1}
    \int_{-\sqrt{1 - x^2}}^{\sqrt{1 - x^2}}
    \frac{\sqrt{1 - x^2 - y^2}}{\pi}
    \,
    \dd y
    \,
    \dd x
    \nonumber \\[0.5em]
    &= \frac{2}{3}
    \quad,
\end{proof}
%
which is precisely the geometric albedo of a Lambert
sphere \citep[see, e.g.][]{Seager2010}.

>>>>

As in \citet{Luger2019}, we compute fluxes by first expanding the surface
in terms of spherical harmonics.
While in \citet{Luger2019} we expanded the emissivity of the surface, here
we instead expand the spherical albedo $A$.
Specifically, if $\bvec{y}$ is the vector
of spherical harmonic coefficients describing the albedo anywhere on
the surface and
$\by$ is the spherical harmonics basis (Equation~\ref{eq:by}),
the albedo $A$ at a point $(x, y)$ on the sky-projected disk of the body
is given by the dot product
%
\begin{align}
    \label{eq:albedo}
    A(x, y) = \by^\top (x, y) \, \bvec{y}
    \quad.
\end{align}
%
The flux measured from this body is the surface integral over the visible
portion of the body's projected disk of the albedo $A$
times the illumination profile $\mathcal{I}$ of the surface, given by
Lambert's law as
%
\begin{align}
    \label{eq:LambertsLaw}
    \mathcal{I}(\vartheta_i) = \mathcal{I}_0 \, \text{max}\big( 0, \cos\vartheta_i \big)
    \quad,
\end{align}
%
where $\mathcal{I}_0$ is the peak illumination (at the sub-source point)
and $\vartheta_i$ is the angle between the incident radiation and the surface
normal.
%

@@

Our last sample light curve is Figure~\ref{fig:earthsun}, which shows
a secondary eclipse light curve of the Earth as it is occulted by the
Sun. The model for the Earth is the same as above, and the observer is
now close to the ecliptic, but slightly misaligned so that the
Earth is occulted behind a solar latitude of $30^\circ$ (i.e., at
a solar impact parameter of $0.5$). The observation takes place at the
June solstice, so the Earth is tilted by $23.5^\circ$ out of the page.
As before, the top panel shows the light curve in thermal light; this
is similar to the top panel of Figure~13 in \citet{Luger2019}. The orange
dots again correspond to the numerical solution.
%
The center panel shows the same light curve in reflected light.
Because the observation occurs very close to full phase, the normalized
light curves look very similar to each other.
The bottom panel shows the difference between the two
(reflected minus thermal), which is only on the order of a few percent.
%
In fact, because
the illumination profile is proportional to the cosine of the viewing
angle, $\upmu$, and the reflection is assumed to be isotropic, the
(normalized) secondary eclipse light curve in reflected light is to
good approximation equal
to a limb-darkened thermal occultation light curve with
linear limb darkening coefficient $u_1 = 1$. As we will see later, for very
close-in planets this approximation breaks down, since the illumination
phases at secondary eclipse ingress and egress are sufficiently different
from full phase.

@@

@@

The algorithm discussed above and derived in the Appendix computes light
curves in the limit that (1) the body is illuminated by a point source
and (2) the body is an ideal Lambertian scatterer. Both of these
assumptions can be relaxed within \starry, and below we discuss modifications
to the code to allow for this.

\subsection{Extended illumination source}
\label{sec:extended}

In the limit that the angular size of the star as seen from the planet is
small, the illumination profile on the surface of the planet will
decrease as the cosine of the angle between the surface normal and the star,
reaching zero at the day/night terminator, an angle $\nicefrac{\pi}{2}$
away from the sub-stellar point. However, if the star is sufficiently large
and the planet is sufficiently close-in, rays originating from near the limb
of the star will reach points on the planet surface beyond this angle.
It is straightforward to show that, provided the stellar radius $R_\star$
is larger than the planet radius $R_p$, the angular extent of the true
day/night terminator past $\nicefrac{\pi}{2}$ is given by
%
\begin{proof}{}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{1 - \nicefrac{R_p}{R_\star}}{\nicefrac{a}{R_\star}} \right)
\end{proof}
%
where $a$ the semi-major axis of the orbit (where we implictly assume
the eccentricity is zero).
For planets like the Earth, this quantity is only about $0.26^\circ$,
resulting in a negligible effect on the planet's light curve. However,
for very close-in planets, the effect can be significant. For instance,
the hot Jupiter Kelt-9b has $\nicefrac{R_p}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$
planet \citep{Wong2019}. Asssuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020},
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of that planet. Figure~\ref{fig:extended} shows the illumination profile
of this planet in a Mollweide projection, with the sub-stellar point at
the center, for the point source approximation (left) and accounting for
the finite size of the star (right). In addition to the
displaced day/night terminator, the main difference between
the two profiles is the sub-stellar intensity, which is significantly higher
in the extended source case. This is due to the simple fact that in the point
source case the illumination source is placed at the center of the star,
which is one stellar radius farther from the planet than the point closest
to the planet (the sub-planetary point) in the extended source case.
Once accounting for this difference, the fractional change in the intensity
on the planet away from the sub-stellar point is similar in both cases,
and the intensity anywhere beyond $\nicefrac{\pi}{2}$ is less than one-tenth
the peak value.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/extended.pdf}
        \oscaption{extended}{%
            Normalized surface intensity on Kelt-9b viewed in a
            Mollweide projection assuming a point illumination source
            (left) and accounting for the finite extent of the star
            (right). The day/night terminator extends about $17^\circ$
            past where it is in the point source case. The sub-stellar
            intensity is higher in the extended source case because the
            sub-planet point on the star is
            closer to the planet than in the case where the star is a point
            source located at the center of the star.
            \label{fig:extended}
        }
    \end{centering}
\end{figure}

The illumination profile in the extended source case may be computed as the
two-dimensional integral of the point source illumination profile over
the visible portion of the stellar disk. While this integral may in theory be
computed analytically, the resulting profile on the planet surface
will not in general be exactly expressible in terms of spherical harmonics, a
necessary condition for the \starry algorithm.
For simplicity, we therefore compute the illumination profile for
extended sources by averaging the contribution of \textsf{source\_npts} point
sources uniformly distributed across the projected disk corresponding to
the portion of the stellar surface
visible from the planet, where \textsf{source\_npts} is a user-supplied
value. In the limit $R_p \ll R_\star$, this is a spherical cap centered at
the sub-planet point with radius $R_\star \cos\tau$. In
Figure~\ref{fig:extended} we set
\textsf{source\_npts} = 300, but in practice we find that ${\sim}30$
points are sufficient for even the most extreme cases such as Kelt-9b.
%
Note, importantly, that while this method allows one to account for the
effect of stellar limb darkening on the illumination profile of the planet,
this has not been implemented in \starry.

@@

Figure~\ref{fig:kelt9b} shows the practical implications of the finite
stellar size for the phase curve of Kelt-9b. The left panel shows the
reflected light phase curve of the planet in parts per million, assuming
a spherical albedo of 0.2, for the point source approximation
(blue) and including the effect of the extended source with
\textsf{source\_npts} = 300 (purple). The increased
illumination at the sub-stellar point results in a ${\sim}30$ increase
in the value of the phase curve close to full phase ($\pm\nicefrac{1}{2}$).
Close to a phase of zero, the extended source results in decreased
flux, since the portion of the star illuminating the planet (the region
close to the limb) is slightly
\emph{farther} away, by a factor of
$\sqrt{1 + (\nicefrac{a}{R_\star})^{-2}}$. This results in a steeper
phase curve, which can be seen in the right panel, where the light curves
have been normalized to their maximum value.

\begin{figure}[p!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/kelt9b_eclipse.pdf}
        \oscaption{kelt9b}{%
            Normalized reflected light secondary eclipse model for Kelt-9b,
            assuming a spherical albedo $A = 0.2$. Blue is the \starry model
            assuming a point source illumination; purple accounts for
            the finite size of the star. The orange curves show
            approximate models computed using the \citet{MandelAgol2002}
            model: a sphere of uniform intensity (solid orange)
            and a sphere whose intensity falls as the cosine of the
            viewing angle from the center of the planet disk
            (dashed orange).
            Because Kelt-9b is so close to its host star, the illumination
            phase changes significantly from ingress to egress, and neither
            approximation accurately captures the behavior of the light curve.
            On the other hand, the point source approximation agrees well with
            the extended source solution, modulo the difference in
            the depth (see Figure~\ref{fig:kelt9b}).
            \label{fig:kelt9b_eclipse}
        }
    \end{centering}
\end{figure}

While modeling the extended size of the star is essential to
getting the shape of the phase curve correct, the same is not true
for secondary eclipse. Figure~\ref{fig:kelt9b_eclipse} shows
the normalized secondary eclipse model for Kelt-9b under the point
source approximation (blue) and the extended source model (purple).
Neglecting the fact that the \emph{depth} of secondary eclipse is
significantly different between the two models (see the left panel
of Figure~\ref{fig:kelt9b}), the difference in \emph{shape} between
the two curves is almost negligible.
%
For reference, the figure shows
two additional models one might consider using to fit a secondary
eclipse light curve: a uniform (unilluminated) disk (solid orange)
and a disk whose intensity falls as $\upmu$, the cosine of the
viewing angle (dashed orange). Both can be computed using the
classical \citet{MandelAgol2002} transit model; the latter corresponds
to a linearly limb-darkened sphere and is functionally equivalent to
a Lambertian sphere seen at full phase.
However, neither approximation is particularly good, since Kelt-9b
changes illumination phase significantly from ingress to egress
owing to its proximity to the star.

\subsection{Non-Lambertian scatterers}
\label{sec:nonlambertian}

The second assumption we now seek to relax is that of Lambertian
scattering. A perfect Lambert sphere reflects light isotropically, so
the measured intensity at a point on the surface is strictly
proportional to the product of the cosine of the angle of incidence
and the cosine of the viewing angle. While this is convenient
from a modeling standpoint, it is hardly ever true in practice.
For planets and moons in particular, there is often a strong phase
dependence in the scattering. Rayleigh scattering in planetary
atmospheres is preferentially in the forward/backward direction,
while clouds and oceans can contribute strong specular reflection.
Moreover, rough surfaces can have complex scattering behavior due
to changes in the orientation of the surface normal on small scales
and effects such as multiple reflections and self-shading.

In principle, any of these processes can be accounted for in
the \starry algorithm by modiying the illumination matrix
$\mathbf{I}$ (Equation~\ref{eq:Imat}).
\xxx{TODO} \citep{OrenNayar1994}.

@@

@@

\section{Discussion}
\label{sec:discussion}

\subsection{Information theory}
\label{sec:information}

\xxx{TODO.}

\section{Conclusions}
\label{sec:conclusionss}

\xxx{TODO.}

@@

Second, we must explicitly model the illumination of the body. We assume the
body is illuminated by a point-like source whose flux measured by the observer
is unity. In this case, the observed intensity at any
point on the surface is proportional to the cosine of the angle $\vartheta_i$ between
the incident light and the surface normal. Points for which
$\vartheta_i \ge \nicefrac{\pi}{2}$ are unilluminated and
therefore have an intensity of zero.
%
If the point-like illumination source is placed at sky coordinates
$(x_s, y_s, z_s)$ in units of the radius of the illuminated body,
the day/night terminator on the body is a half-ellipse
of semi-major axis unity that is fully described by its (signed) semi-minor
axis,
%
\begin{proof}{}
    \label{eq:b}
    b = -\frac{z_s}{r_s}
    \quad,
\end{proof}
%
where $r_s = \sqrt{x_s^2 + y_s^2 + z_s^2}$ is the distance to the source,
%
and the angle $\theta$ by which its semi-major axis is rotated away from the
$+x$-axis (a frame-dependent quantity, defined below).
Given this formulation, and assuming that
$r_s \gg 1$,
it is straightforward to show that the illumination
$\mathcal{I}$ at a point $(x, y)$ on the projected disk of the body is given
by the function
%
\begin{proof}{}
    \label{eq:illum}
    \mathcal{I}(b, \theta, r_s; x, y)&=
    \text{max}\bigg( 0, I(b, \theta, r_s; x, y) \bigg)
\end{proof}
%
where
%
\begin{proof}{}
    \label{eq:illum_poly}
    I(b, \theta, r_s; x, y) &= \frac{1}{\pi r_s^2}
    \bigg(
    -b_c\sin\theta x + b_c\cos\theta y - bz(x, y)
    \bigg)
\end{proof}
%
with $b_c \equiv \sqrt{1 - b^2}$ and $z(x, y) = \sqrt{1 - x^2 - y^2}$.
A quick note on the normalization of Equation~(\ref{eq:illum_poly}):
if we place the illumination source along the $+z$-axis at $(0, 0, 1)$,
the body is seen at full phase, so $b = -1$, $b_c = 0$, and
%
\begin{proof}{}
    \mathcal{I}_\text{full}(x, y) = \frac{\sqrt{1 - x^2 - y^2}}{\pi}
    \quad.
\end{proof}
%
Integrating this over the full disk, we obtain the reflected flux measured
by the observer if the body has unit spherical albedo,
%
\begin{proof}{}
    \mathcal{f}_\text{full} &=
    \int_{-1}^{1}
    \int_{-\sqrt{1 - x^2}}^{\sqrt{1 - x^2}}
    \frac{\sqrt{1 - x^2 - y^2}}{\pi}
    \,
    \dd y
    \,
    \dd x
    \nonumber \\[0.5em]
    &= \frac{2}{3}
    \quad,
\end{proof}
%
which is precisely the geometric albedo of a Lambert
sphere \citep[see, e.g.][]{Seager2010}.
